Pituitary prolactin drives transcriptional changes in the reproductive axis of parenting pigeons

### 2020-03-07 docs/manuscript/elife-v1.tex
mew figs

% !TeX root = RJwrapper.tex
\title{Internal physiology and external cues drive changes in reproductive axis
of parenting pigeons}
\author{by Rayna M Harris \footnote{equal contribtion}, Suzanne H. Austin \footnote{equal contribtion}, Andrew S. Lang, Victoria S. Farrar, April Booth, Tanner Feustel, Matthew D. MacManes, Rebecca M. Calisi \footnote{corresponding author}}

@@

% Any extra LaTeX you need in the preamble

In species that provide parental care, successful rearing of offspring involves a shift from aggressive and sexual behaviors to more caring and nurturing ones.  We hypothesize that these shifts in behavior are related to changes in gene expression that are controlled by internal mechanisms rather than by the external cues.  We found that gene expression in the hypothalamus is fairly insensitive to the natural transition between egg and nestling stages; however, in the pituitary and gonads, gene expression showed much more plasticity over the course of reproduction. We show evidence that natural transitions in reproductive and parental behavior by mediated internal clock controlling gene expression in the pitary and gonads but that the hypothalamus is very sensitive to external cues that signal a distruption in the natural parental care cycle. This research provides a deeper understanding of the interplay of genes, hormones, and parental care and has important implications for understanding molecular processes that underlie behavior transitions.

@@

@@

Our results suggest that tissue is the greatest source of variation in
the dataset, with hypothalamic, pituitary, and gonadal samples all
forming distinct clusters, with only a few pituitary and gonad samples
falling outside the 95\% confidence interval (Fig. \ref{figure:fig1}b).
Sex-differences are most prounounced in the gonads with male and female
samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment
accounts for much less of the variation than tissue and sex, and even
after removing sex as a variatble, treatment effects in all three
tissues are hard to visualize simultatenously (Fig.
\ref{figure:fig1}d,e), so we next looked at each tissue and sex
separately. In the hypothalamus, control samples form tight clusters
that stands out from all other timepoints that are largely overlapping
in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest
treatment effect with one cluster of the ``earlier'' parental timepoints
(nest-building, lay, incubation days 3 and 9), a second cluster of
``later'' parental timpoints (incubation day 17, hatch, and nestling
care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in
the gonads, the control timepoints appear to be much less variable that
the parental timepoints, but there is no clear separation of gonadal
samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE
analysis reveals broad patterns across all levels of the experiment,
additional tools are needed it hone in on the specific differences
associated with parental care in each tissue.

Differential gene expression analysis can shed insight into specific
differences between treatment groups and thus deepen our understand of
the transitions between reproduction and parental care behaviors. We
used DESEq2 to identify genes whose expression changed significantly
between sequential timepoints in each tissue and sex separately (Fig.
\ref{figure:fig2}a).

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig1-1}
  \caption{Broad patterns of variation in hypothalamic, pituitary, and gonadal gene expression in male and female rock doves across reproductive and parental transitions. a) Experimental design. b) t-SNE plot highlighting tissue differences in gene expression. Each dot represents 1 sample and is colored by the source tissue (hypothalamus: orange, pituitary: teal, gonads: purple). c) t-SNE plot highlighting sex differences in tissue and sex, as shown in light grey (females) and dark grey (males). t-SNE plots highlight treatment effects in females (d), males (e), the hypothalamus (h), pituitary (i), and gonads (j), with samples colored by parental stage (control: red or , nest-building: orangse, lay: yellow; inc.d3: light green, inc.d9: green, inc.d17: dark green, hatch: dark blue, n5: blue, and n9: light blue.}
  \label{figure:fig1}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig4-1}
  \caption{Another figure}
  \label{figure:fig4}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig5-1}
  \caption{Another another figure}
  \label{figure:fig5}
\end{figure}

@@

Results in context to literature.

Significance.

Limitations.

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig3}
  \caption{R shiny, image 1}
  \label{figure:supplfig3}
\end{figure}

@@

\hypertarget{animal-care}{%
\subsubsection{Animal Care}\label{animal-care}}

@@

\hypertarget{experimental-design-and-tissue-processing}{%
\subsubsection{Experimental design and tissue
processing}\label{experimental-design-and-tissue-processing}}

@@

\hypertarget{rna-sequencing-analysis}{%
\subsubsection{RNA sequencing analysis}\label{rna-sequencing-analysis}}

@@

\hypertarget{data-availablility}{%
\subsubsection{Data availablility}\label{data-availablility}}

@@

\address{%
Rayna M Harris \footnote{equal contribtion}\\
UC Davis\\
\\
}

\address{%
Suzanne H. Austin \footnote{equal contribtion}\\
UC Davis\\
\\
}

\address{%
Andrew S. Lang\\
UNH\\
\\
}

\address{%
Victoria S. Farrar\\
UC Davis\\
\\
}

\address{%
April Booth\\
UC Davis\\
\\
}

\address{%
Tanner Feustel\\
UC Davis\\
\\
}

\address{%
Matthew D. MacManes\\
UNH\\
\\
}

\address{%
Rebecca M. Calisi \footnote{corresponding author}\\
UC Davis\\
\\
}
\href{mailto:author7@work}{\nolinkurl{author7@work}}

>>>>

% !TeX root = RJwrapper.tex
\title{Pituitary prolactin drives transcriptional changes in the reproductive
axis of parenting pigeons}
\author{by Rayna M Harris \footnote{equal contribtion}, Suzanne H. Austin \footnote{equal contribtion, currently at the
  University of Oregon}, Andrew S. Lang, Victoria S. Farrar, April Booth, Tanner Feustel, Matthew D. MacManes, Rebecca M. Calisi \footnote{corresponding author}}

@@

\abstract{%
In species that provide parental care, successful rearing of offspring
involves a shift from aggressive and sexual behaviors to more caring and
nurturing ones. We hypothesize that these shifts in behavior are related
to changes in gene expression that are controlled by internal mechanisms
rather than by the external cues. We found that gene expression in the
hypothalamus is fairly insensitive to the natural transition between egg
and nestling stages; however, in the pituitary and gonads, gene
expression showed much more plasticity over the course of reproduction.
We show evidence that natural transitions in reproductive and parental
behavior by mediated internal clock controlling gene expression in the
pitary and gonads but that the hypothalamus is very sensitive to
external cues that signal a distruption in the natural parental care
cycle. This research provides a deeper understanding of the interplay of
genes, hormones, and parental care and has important implications for
understanding molecular processes that underlie behavior transitions.
}

@@

\emph{Characterization of parental care}

@@

Our results suggest that tissue is the greatest source of variation in
the dataset, with hypothalamic, pituitary, and gonadal samples all
forming distinct clusters, with only a few pituitary and gonad samples
falling outside the 95\% confidence interval (Fig. \ref{figure:fig1}b).
Sex-differences are most prounounced in the gonads with male and female
samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment
accounts for much less of the variation than tissue and sex, and even
after removing sex as a variatble, treatment effects in all three
tissues are hard to visualize simultatenously (Fig.
\ref{figure:fig1}d,e), so we next looked at each tissue and sex
separately. In the hypothalamus, control samples form tight clusters
that stands out from all other timepoints that are largely overlapping
in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest
treatment effect with one cluster of the ``earlier'' parental timepoints
(nest-building, lay, incubation days 3 and 9), a second cluster of
``later'' parental timpoints (incubation day 17, hatch, and nestling
care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in
the gonads, the control timepoints appear to be much less variable that
the parental timepoints, but there is no clear separation of gonadal
samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE
analysis reveals broad patterns across all levels of the experiment,
additional tools are needed it hone in on the specific differences
associated with parental care in each tissue. Differential gene
expression analysis can shed insight into specific differences between
treatment groups and thus deepen our understand of the transitions
between reproduction and parental care behaviors. We used DESEq2 to
identify genes whose expression changed significantly between sequential
timepoints in each tissue and sex separately (Fig. \ref{figure:fig2}a).

@@

\emph{Internal physiology versus external stimuli hypothesis}

We show that samples from the pituitary on incubation day 17 cluster
more closely with hatch and nestling care day 5 samples than with
incubation days 3 and 9. This suggest that, in the pituitary, internal
mechanisms that prepare for chick arrival have the greatest effect on
gene expression rather than suggesting that gene expression responds to
changes in the external enviornament.

Blah \ref{figure:fig3}, blah, \ref{figure:fig4} blah.

\emph{Potential implications for cancer research}

These results confirmed prolactin (PRL) gene expression fluctuates
throughout parental care in a manner that is consistent with its role in
promoting lactation and maintaining parental behaviors. Interestingly,
we identified about 100 genes whose expression was correlated with PRL,
including (BRCA1, which is associated with breast cancer). Additionally,
we identified thousands of genes whose expression changed over the
course of parental care. Exploring the relationship between PRL and
other differentially expressed genes could provide important insights
into the \texttt{symphony} of gene expression that regulates behavior at
the organismal and cellular level.

\emph{Software for data exploration and hypothesis testing}

@@

\emph{Results in context to literature. }

blah blah balh

\emph{Significance. }

blah blah balh

\emph{Limitations.}

blah blah balh

@@

\emph{Animal Care}

@@

\emph{Experimental design and tissue processing}

@@

\emph{RNA sequencing analysis}

@@

\emph{Data availablility}

@@

\newpage

\newpage

\newpage

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig3-1}
  \caption{Another figure}
  \label{figure:fig3}
\end{figure}

\newpage

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig4-1}
  \caption{Another another figure}
  \label{figure:fig4}
\end{figure}

\newpage

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig3}
  \caption{R shiny, image 1}
  \label{figure:supplfig3}
\end{figure}

\newpage

\hypertarget{author-affiliations}{%
\subsection{Author affiliations}\label{author-affiliations}}

\address{%
Rayna M Harris \footnote{equal contribtion}\\
University of California, Davis\\
\\
}

\address{%
Suzanne H. Austin \footnote{equal contribtion, currently at the
  University of Oregon}\\
University of California, Davis\\
\\
}

\address{%
Andrew S. Lang\\
University of New Hampshire\\
\\
}

\address{%
Victoria S. Farrar\\
University of California, Davis\\
\\
}

\address{%
April Booth\\
University of California, Davis\\
\\
}

\address{%
Tanner Feustel\\
University of California, Davis\\
\\
}

\address{%
Matthew D. MacManes\\
University of New Hampshire\\
\\
}

\address{%
Rebecca M. Calisi \footnote{corresponding author}\\
University of California, Davis\\
\\
}
\href{mailto:rmcalisi@ucdavis.edu}{\nolinkurl{rmcalisi@ucdavis.edu}}
